---
title: "CosMx UAP1 SSc - immune cell annotation by PopV with GSE195452 skin dataset"
author: "Yi-Nan Li"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_float: yes
    df_print: paged
    wrap: 72
---

## Load data (USER INPUT REQUIRED)

Load the annotated data from Alex with BANKSY clustering and GSE195452 skin dataset from scratch.

```{r data}
### Preparation for PopV inputs ###

library(SpatialExperiment)
library(tidyverse)
library(rmarkdown)
spe <- readr::read_rds("~/CosMx/2024-07-16 _CosMxSScSkin_BANKSY_spe_CN.rds")

## Filter BANKSY annotated immune cells
spe <- spe[,spe$lv1.anno_bank == 'Immune']


# ## load GSE195452 skin immune cell from raw count matrices
# dat.dir <- '~/data/scRNAseq/GSE195452_Amit-LGR5/'
# GSE195452.SRA.meta <- read.csv("~/data/scRNAseq/GSE195452_Amit-LGR5/SraRunTable.txt")
# GSE195452.anno <- read.csv("~/data/scRNAseq/GSE195452_Amit-LGR5/GSE195452_Cell_metadata_v26_anno.txt", sep = "\t")
# GSE195452.skin.samples <- GSE195452.SRA.meta %>% filter(Tissue == "Skin" & Assay.Type == "RNA-Seq") %>%
#   pull(Sample.Name)
# sample.to.analysis <- list.files(paste0(dat.dir,"/GSE195452_RAW"))[grepl(paste(GSE195452.skin.samples, collapse='|'),
#                                                                         list.files(paste0(dat.dir,"/GSE195452_RAW")))]
# ## Load the count matrix
# library(scuttle)
# mat <- readSparseCounts(paste0(dat.dir,"GSE195452_RAW/",sample.to.analysis[1]))
# GSE195452.rawcount <- data.frame(row.names = row.names(mat)) %>% as.matrix() %>% Matrix::Matrix(sparse = TRUE)
# for (i in sample.to.analysis[-1]) {
#   tab <- readSparseCounts(paste0(dat.dir,"GSE195452_RAW/", i))
#   if (identical(row.names(GSE195452.rawcount), row.names(tab))) {
#     cat(paste0("Reading data from ", i, " (", which(i == sample.to.analysis), "/", length(sample.to.analysis), ")\n"))
#     GSE195452.rawcount <- cbind(GSE195452.rawcount, tab) 
#   }
#   else{
#     warning(paste0("Row names do not match for ", i, ". Sample skipped!! \n"))
#   }
# }
# ref_sce <- SingleCellExperiment(assay = list(counts = GSE195452.rawcount), 
#                                 colData = GSE195452.anno[colnames(GSE195452.rawcount),])
# 
# ref_sce$lv1_anno <- ref_sce$lv1.anno <- case_when(startsWith(ref_sce$annotation, "B") ~ "Bcell",
#                                      startsWith(ref_sce$annotation, "Fibro") ~ "Fibroblast",
#                                      startsWith(ref_sce$annotation, "T") ~ "Tcell",
#                                      startsWith(ref_sce$annotation, "sT") ~ "Tcell",
#                                      startsWith(ref_sce$annotation, "Vascular") ~ "Endothelial",
#                                      startsWith(ref_sce$annotation, "Lymphatic") ~ "Endothelial",
#                                      startsWith(ref_sce$annotation, "NK") ~ "NK",
#                                      startsWith(ref_sce$annotation, "Peri") ~ "Pericyte",
#                                      startsWith(ref_sce$annotation, "KRT") ~ "Epithelial",
#                                      startsWith(ref_sce$annotation, "Mf") ~ "Macrophage",
#                                      startsWith(ref_sce$annotation, "Mo") ~ "Monocyte",
#                                      startsWith(ref_sce$annotation, "M_") ~ "Monocyte",
#                                      startsWith(ref_sce$annotation, "Mast") ~ "Mastcell",
#                                      startsWith(ref_sce$annotation, "DC") ~ "DC",
#                                      startsWith(ref_sce$annotation, "NK") ~ "NK",
#                                      .default = "other")
# 
# # import patient info
# sample_id <- data.frame(sample_id = str_split_i(sample.to.analysis, "_", 1), Amp_batch_ID = str_split_i(sample.to.analysis, "_", 2))
# sample_id$Amp_batch_ID <- gsub(".txt.gz", "", sample_id$Amp_batch_ID)
# sum(duplicated(sample_id$Amp_batch_ID)) # check if amp_id is duplicated
# patient_id <- left_join(GSE195452.SRA.meta, sample_id, by = c(`Sample.Name` = "sample_id")) %>% select(PATIENT_ID, `Sample.Name`, Amp_batch_ID, selection_marker)
# patient_ids <- pull(patient_id, PATIENT_ID) %>% `names<-`(patient_id$Amp_batch_ID)
# 
# ref_sce$patient_id <- patient_ids[ref_sce$Amp_batch_ID] %>% unname()
# # pat_info <- readxl::read_xlsx(paste0(dat.dir, "GSE195452_PatientInfo_mmc1.xlsx"), skip = 2) %>% as.data.frame() 
# # rownames(pat_info) <- pat_info$PID
# 
# # save sce for future use
# readr::write_rds(ref_sce, '~/data/scRNAseq/GSE195452_AllSkinCell_sce.rds')

ref_sce <- readr::read_rds('~/data/scRNAseq/GSE195452_AllSkinCell_sce.rds')
ref_sce$annotation <- factor(ref_sce$annotation)
ref_sce$lv1_anno <- factor(ref_sce$lv1_anno)

## remove problematic dots ##
cols_to_keep <- colnames(colData(ref_sce))[!grepl("\\.", colnames(colData(ref_sce)))]
colData(ref_sce) <- colData(ref_sce)[, cols_to_keep, drop = FALSE]
  
ref_sce <- ref_sce[,ref_sce$lv1_anno %in% c('Tcell', 'NK', 'Macrophage', 'Monocyte', 'Mastcell', 'DC', 'Bcell') & ref_sce$Number_of_cells == 1]

## setting up the PopV conda env

tryCatch({
  reticulate::use_condaenv("PopV")
  reticulate::py_config()
}, error = function(e) {
  message("Error setting up Python environment: ", e$message)
  quit(status = 1)
})

## Convert spe to anndata
library(zellkonverter)
# open issue for sce dev team
# see https://github.com/theislab/zellkonverter/issues/113
mcols(colPairs(spe)[[1]])$value <- 1
# remove table in table in spe that might give errors when writing AnnData
for (i in colnames(colData(spe))) {
  if (class(spe[[i]]) == "DFrame") {
    spe[[i]] <- NULL
  }
}
colData(spe) <- cbind(colData(spe), spatialCoords(spe))

adata <- SCE2AnnData(spe)
ref_adata <- SCE2AnnData(ref_sce)

```

## Load python modules and data transfer from R - **USER INPUT REQUIRED**

```{python popv-setup}
import anndata as ad
import popv
import numpy as np
import scanpy as sc
import scipy.sparse as sp


### Select the needed metadata ###
## create clean metadata to reduce errors
columns_to_keep = ['cell', 'donor', 'block', 'sample_id']

query_adata = r.adata
query_adata.obs['query_labels'] = 'unknown'
query_adata.obs['query_labels'] = query_adata.obs['query_labels'].astype('category')
# Convert the dense matrix to a sparse matrix (CSR format)
query_adata.X = sp.csr_matrix(query_adata.X)

query_adata.obs = query_adata.obs[columns_to_keep]

ref_adata = r.ref_adata

```


## Start PopV pipeline

### Setting Up Annotation Parameters - **USER INPUT REQUIRED**
Here is where you set the parameters for the automated annotation.

Arguments:

* **tissue**: Tabula Sapiens tissue to annotate your data with. Available tissues: ["Bladder", "Blood", "Bone_Marrow", "Kidney", "Large_Intestine", "Lung","Lymph_Node", "Pancreas", "Small_Intestine", "Spleen", "Thymus","Trachea", "Vasculature"]
* **save_location**: location to save results to. By default will save to a folder named annotation_results. It is highly recommended you provide a Google Drive folder here.
* **query_batch_key**: key in query_adata.obs for batch correction. Set to None for no batch correction.
* **methods**: these are the methods to run. By default, will run all methods.

Lesser used parameters

* **query_labels_key**: scANVI has the option to use labeled cells in the query dataset during training. To use some prelabeled cells from the query dataset, set query_labels_key to the corresponding key in query_adata.obs
* **unknown_celltype_label**: If query_labels_key is not None, will treat everything not labeled unknown_celltype_label as a labeled cell

```{python popv-parm}
""" 
tissue options: 
['Bladder','Blood','Bone_Marrow','Fat',
'Heart','Kidney','Large_Intestine','Liver',
'Lung','Lymph_Node','Mammary','Muscle',
'Pancreas','Prostate','Salivary Gland',
'Skin','Small_Intestine','Spleen',
'Thymus','Trachea','Vasculature']
"""

### So far automatic download TS datsets did not work ###
tissue = 'Skin'
save_folder = 'popv_results'
### set up the output h5ad path ##
out_path = '/mnt/data/home/yinanli/CosMx/PopVannotation/CosMxUAP1SSc_Immun_query_and_GSE195452_popv.h5ad'

### Put the key for batch correction in query dataset ###
query_batch_key = 'block'

### Set the methods ### 
# bbknn, scanorama, and rf performed bad in previouse run
# methods = None # set to None for using all the methods available
methods = ['celltypist', 'knn_on_harmony','knn_on_scvi', 'scanvi', 'svm']
# methods = ['celltypist', 'knn_on_bbknn', 'knn_on_harmony', 'knn_on_scanorama', 'knn_on_scvi', 'onclass', 'rf', 'scanvi', 'svm']

# Lesser used parameters
query_labels_key='query_labels'
unknown_celltype_label='unknown'


from scanpy._utils import check_nonnegative_integers
          
assert check_nonnegative_integers(query_adata.X) == True, 'Make sure query_adata.X contains raw_counts'
```

### Setup reference data (USER INPUT REQUIRED)

```{python ref-setup}
### Put the annotaion and parameter for batch correction here ###
ref_labels_key = "annotation"
ref_batch_key = "patient_id"

assert check_nonnegative_integers(ref_adata.X) == True, 'Make sure ref_adata.X contains raw_counts'

min_celltype_size = np.min(ref_adata.obs.groupby(ref_labels_key).size())
n_samples_per_label = np.max((min_celltype_size, 500))
```

### Preprocess query with ref dataset

```{python query-pp}

from popv.preprocessing import Process_Query

adata = Process_Query(
    query_adata,
    ref_adata,
    query_labels_key=None,
    query_batch_key=query_batch_key,
    ref_labels_key=ref_labels_key,
    ref_batch_key=ref_batch_key,
    unknown_celltype_label=unknown_celltype_label,
    cl_obo_folder = False,
    # cl_obo_folder="./PopV/resources/ontology/",
    prediction_mode="retrain",  # 'fast' mode gives fast results (does not include BBKNN and Scanorama and makes more inaccurate predictions)
    n_samples_per_label=n_samples_per_label,
    accelerator="cpu",
    compute_embedding=True,
    hvg=None,
    save_path_trained_models = '/mnt/data/home/yinanli/CosMx/GSE195452_Immune_models',
).adata
```


```{python popv-anno}
from popv.annotation import annotate_data

annotate_data(adata, methods=methods,save_path="/mnt/data/home/yinanli/CosMx/PopVannotation/Immune_popv_output")

#Optional: save the full anndata will all objects
adata.write(out_path)

import pandas as pd
# adata = sc.read_h5ad(out_path)
cell_types = pd.unique(adata.obs[[i for i in adata.uns['prediction_keys']] + ['popv_parent']].values.ravel('K'))
palette = sc.plotting.palettes.default_102
celltype_colors = {i: j for i, j in zip(list(cell_types), palette)}
```

## PopV visulization

```{r viz, fig.show='hold'}
library(zellkonverter)
library(dittoSeq)
library(dplyr)
library(reticulate)


popv_sce <- readH5AD(py$out_path, verbose = T, reader = 'python')
readr::write_rds(popv_sce, gsub('.h5ad', '_sce.rds', py$out_path))
# popv_seurat <- Seurat::as.Seurat(popv_sce) %>% Seurat::UpdateSeuratObject()
# readr::write_rds(popv_seurat, gsub('.h5ad', '_seurat.rds', py$out_path))

celltype_color <- py$celltype_colors %>% unlist
# 
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scvi_umap_popv", var = "popv_prediction", opacity = .5, do.label = T, labels.size = 3, color.panel = celltype_color)
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scvi_umap_popv", var = "popv_prediction_score", opacity = .5, do.label = T, labels.size = 2)
dittoDimPlot(popv_sce, reduction.use = "X_scanvi_umap_popv", var = "popv_prediction_score", opacity = .5,
             do.label = T, labels.size = 2)
dittoDimPlot(popv_sce, reduction.use = "X_scvi_umap_popv", var = "X_dataset", opacity = .5,
             do.label = F, labels.size = 2, order = 'increasing')
dittoDimPlot(popv_sce, reduction.use = "X_scanvi_umap_popv", var = "X_dataset", opacity = .5,
             do.label = F, labels.size = 2, order = 'increasing')
dittoDimPlot(popv_sce, reduction.use = "X_umap_scanorma_popv", var = "X_dataset", opacity = .5,
             do.label = F, labels.size = 2, order = 'increasing')
dittoDimPlot(popv_sce, reduction.use = "X_bbknn_umap_popv", var = "X_dataset", opacity = .5,
             do.label = F, labels.size = 2, order = 'increasing')
dittoDimPlot(popv_sce, reduction.use = "X_umap_harmony_popv", var = "X_dataset", opacity = .5,
             do.label = F, labels.size = 2, order = 'increasing')

dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_umap_harmony_popv", var = "popv_prediction", opacity = .3,
             do.label = T, labels.size = 3, color.panel = celltype_color)
dittoDimPlot(popv_sce, reduction.use = "X_scvi_umap_popv", var = "popv_prediction", opacity = .3,
             do.label = T, labels.size = 3, color.panel = celltype_color)
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scanvi_umap_popv", var = "popv_prediction", opacity = .3,
             do.label = T, labels.size = 3, color.panel = celltype_color)

dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'ref'], reduction.use = "X_scvi_umap_popv", var = "popv_parent", opacity = .1,
             do.label = T, labels.size = 3, color.panel = celltype_color)


```

## Session information

```{r session}
sessioninfo::session_info()

```

