---
title: "CosMx UAP1 SSc - immune cell annotation by PopV with GSE195452 skin dataset"
author: "Yi-Nan Li"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_float: yes
    df_print: paged
    wrap: 72
    code_folding: hide
---

```{r ts}
# setup timestamp
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()
```

This document was created with the time stamp *`r ts`*. <br>

## Load data - **USER INPUT REQUIRED**

Put the path to the input data. The script currently expecting the data in `SingleCellExperiment` format. <br>
* **QUERY_PATH** : the path to the qurery dataset (in rds). Can also be a `SpatialExperiment` object. <br>
* **REF_PATH**: the path to the reference dataset (in rds). <br> 
* **ANNOTATION**: the column where the annotation is saved in the reference dataset in `colData`. <br>
  *Note: don't include "." in the annotation/batch column names*


```{r data, message=FALSE}

### STRAT OF USER INPUTS ###

QUERY_PATH = '~/CosMx/2024-07-16 _CosMxSScSkin_BANKSY_spe_CN.rds'
REF_PATH = '~/data/scRNAseq/GSE195452_AllSkinCell_sce.rds'
ANNOTATION = 'annotation'






library(SpatialExperiment)
library(tidyverse)
library(rmarkdown)
spe <- readr::read_rds(QUERY_PATH)

ref_sce <- readr::read_rds(REF_PATH)
ref_sce[[ANNOTATION]] <- factor(ref_sce[[ANNOTATION]])

## remove problematic dots ##
cols_to_keep <- colnames(colData(ref_sce))[!grepl("\\.", colnames(colData(ref_sce)))]
colData(ref_sce) <- colData(ref_sce)[, cols_to_keep, drop = FALSE]


## setting up the PopV conda env

tryCatch({
  reticulate::use_condaenv("PopV")
  reticulate::py_config()
}, error = function(e) {
  message("Error setting up Python environment: ", e$message)
  quit(status = 1)
})

## Convert spe to anndata
library(zellkonverter)
# open issue for sce dev team
# see https://github.com/theislab/zellkonverter/issues/113
mcols(colPairs(spe)[[1]])$value <- 1
# remove table in table in spe that might give errors when writing AnnData
for (i in colnames(colData(spe))) {
  if (class(spe[[i]]) == "DFrame") {
    spe[[i]] <- NULL
  }
}
colData(spe) <- cbind(colData(spe), spatialCoords(spe))

adata <- SCE2AnnData(spe)
ref_adata <- SCE2AnnData(ref_sce)

```

## Load python modules and data transfer from R - **USER INPUT REQUIRED**

```{python popv-setup}
import anndata as ad
import popv
import numpy as np
import scanpy as sc
import scipy.sparse as sp


### Select the needed metadata in the query dataset ###
## create clean metadata to reduce errors
columns_to_keep = ['cell', 'donor', 'block', 'sample_id']

query_adata = r.adata
query_adata.obs['query_labels'] = 'unknown'
query_adata.obs['query_labels'] = query_adata.obs['query_labels'].astype('category')
# Convert the dense matrix to a sparse matrix (CSR format)
query_adata.X = sp.csr_matrix(query_adata.X)

query_adata.obs = query_adata.obs[columns_to_keep]

ref_adata = r.ref_adata

```


## Start PopV pipeline

### Setting Up Annotation Parameters - **USER INPUT REQUIRED**
Here is where you set the parameters for the automated annotation.

Arguments:

* **tissue**: Tabula Sapiens tissue to annotate your data with. Available tissues: ["Bladder", "Blood", "Bone_Marrow", "Kidney", "Large_Intestine", "Lung","Lymph_Node", "Pancreas", "Small_Intestine", "Spleen", "Thymus","Trachea", "Vasculature"]
* **save_location**: location to save results to. By default will save to a folder named annotation_results. It is highly recommended you provide a Google Drive folder here.
* **query_batch_key**: key in query_adata.obs for batch correction. Set to None for no batch correction.
* **methods**: these are the methods to run. By default, will run all methods.Options: ['celltypist', 'knn_on_bbknn', 'knn_on_harmony', 'knn_on_scanorama', 'knn_on_scvi', 'onclass', 'rf', 'scanvi', 'svm']

Lesser used parameters

* **query_labels_key**: scANVI has the option to use labeled cells in the query dataset during training. To use some prelabeled cells from the query dataset, set query_labels_key to the corresponding key in query_adata.obs
* **unknown_celltype_label**: If query_labels_key is not None, will treat everything not labeled unknown_celltype_label as a labeled cell

```{python popv-parm}
""" 
tissue options: 
['Bladder','Blood','Bone_Marrow','Fat',
'Heart','Kidney','Large_Intestine','Liver',
'Lung','Lymph_Node','Mammary','Muscle',
'Pancreas','Prostate','Salivary Gland',
'Skin','Small_Intestine','Spleen',
'Thymus','Trachea','Vasculature']
"""

### So far automatic download TS datsets did not work ###
tissue = 'Skin'
save_folder = 'popv_results'
### set up the output h5ad path ##
out_path = '/mnt/data/home/yinanli/CosMx/PopVannotation/CosMxUAP1SSc_Fib_GSE195452_popv_' + r.ts +'.h5ad'

### Put the key for batch correction in query dataset ###
query_batch_key = 'block'

### Set the methods ### 
# bbknn, scanorama, and rf performed bad in previouse run
# methods = None # set to None for using all the methods available
methods = None


# Lesser used parameters
query_labels_key='query_labels'
unknown_celltype_label='unknown'


from scanpy._utils import check_nonnegative_integers
          
assert check_nonnegative_integers(query_adata.X) == True, 'Make sure query_adata.X contains raw_counts'
```

### Setup reference data **USER INPUT REQUIRED**

```{python ref-setup}
### Put the annotaion and parameter for batch correction here ###
ref_labels_key = "annotation"
ref_batch_key = "patient_id"

assert check_nonnegative_integers(ref_adata.X) == True, 'Make sure ref_adata.X contains raw_counts'

min_celltype_size = np.min(ref_adata.obs.groupby(ref_labels_key).size())
n_samples_per_label = np.max((min_celltype_size, 500))
```

### Preprocess query with ref dataset

```{python query-pp}

from popv.preprocessing import Process_Query

adata = Process_Query(
    query_adata,
    ref_adata,
    query_labels_key=None,
    query_batch_key=query_batch_key,
    ref_labels_key=ref_labels_key,
    ref_batch_key=ref_batch_key,
    unknown_celltype_label=unknown_celltype_label,
    cl_obo_folder = False,
    # cl_obo_folder="./PopV/resources/ontology/",
    prediction_mode="retrain",  # 'fast' mode gives fast results (does not include BBKNN and Scanorama and makes more inaccurate predictions)
    n_samples_per_label=n_samples_per_label,
    accelerator="cpu",
    compute_embedding=True,
    hvg=None
).adata
```


```{python popv-anno}
from popv.annotation import annotate_data

annotate_data(adata, methods=methods)

#Optional: save the full anndata will all objects
adata.write(out_path)

import pandas as pd
# adata = sc.read_h5ad(out_path)
cell_types = pd.unique(adata.obs[[i for i in adata.uns['prediction_keys']] + ['popv_parent']].values.ravel('K'))
palette = sc.plotting.palettes.default_102
celltype_colors = {i: j for i, j in zip(list(cell_types), palette)}
```

## PopV visulization {.tabset .tabset-pill}

```{r viz, results='asis'}
library(zellkonverter)
library(SingleCellExperiment)
library(dittoSeq)
library(dplyr)
library(reticulate)

popv_sce <- readH5AD(py$out_path, verbose = T, reader = 'python')
readr::write_rds(popv_sce, gsub('.h5ad', '_sce.rds', py$out_path))
# popv_seurat <- Seurat::as.Seurat(popv_sce) %>% Seurat::UpdateSeuratObject()
# readr::write_rds(popv_seurat, gsub('.h5ad', '_seurat.rds', py$out_path))

celltype_color <- py$celltype_colors %>% unlist

cat('\n\n### PopV prediction \n')
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'ref'], reduction.use = "X_scanvi_umap_popv", var = "popv_parent", opacity = .1,
             do.label = T, labels.size = 3, color.panel = celltype_color, main = 'ref_annotation') 
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scanvi_umap_popv", var = "popv_prediction", opacity = .5, do.label = T, labels.size = 3, color.panel = celltype_color, main = 'PopV prediction')
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scanvi_umap_popv", var = "popv_prediction_score", opacity = .5, do.label = T, labels.size = 2, main ='Prediction score')

cat('\n\n### Embeddings \n')
dittoDimPlot(popv_sce, reduction.use = "X_scvi_umap_popv", var = "X_dataset", opacity = .5,
             do.label = F, labels.size = 2, order = 'increasing', main = 'scVI')
dittoDimPlot(popv_sce, reduction.use = "X_scanvi_umap_popv", var = "X_dataset", opacity = .5,
             do.label = F, labels.size = 2, order = 'increasing', main = 'scANVI')
dittoDimPlot(popv_sce, reduction.use = "X_umap_scanorma_popv", var = "X_dataset", opacity = .5,
             do.label = F, labels.size = 2, order = 'increasing', main = 'scanorma')
dittoDimPlot(popv_sce, reduction.use = "X_bbknn_umap_popv", var = "X_dataset", opacity = .5,
             do.label = F, labels.size = 2, order = 'increasing', main = 'BBKNN')
dittoDimPlot(popv_sce, reduction.use = "X_umap_harmony_popv", var = "X_dataset", opacity = .5,
             do.label = F, labels.size = 2, order = 'increasing', main = 'Harmony')


cat('\n\n### scVI \n') 
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'ref'], reduction.use = "X_scvi_umap_popv", var = "popv_parent", opacity = .1,
             do.label = T, labels.size = 3, color.panel = celltype_color, main = 'ref_annotation') 
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scvi_umap_popv", var = "popv_knn_on_scvi_prediction", opacity = .5, do.label = T, labels.size = 3, color.panel = celltype_color)
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scvi_umap_popv", var = "popv_knn_on_scvi_prediction_probabilities", opacity = .5, do.label = T, labels.size = 3)

cat('\n\n### scANVI \n')
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'ref'], reduction.use = "X_scanvi_umap_popv", var = "popv_parent", opacity = .1,
             do.label = T, labels.size = 3, color.panel = celltype_color, main = 'ref_annotation') 
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scanvi_umap_popv", var = "popv_scanvi_prediction", opacity = .5, do.label = T, labels.size = 3, color.panel = celltype_color)
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scanvi_umap_popv", var = "popv_scanvi_prediction_probabilities", opacity = .5, do.label = T, labels.size = 3)

cat('\n\n### Harmony \n')
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'ref'], reduction.use = "X_umap_harmony_popv", var = "popv_parent", opacity = .1,
             do.label = T, labels.size = 3, color.panel = celltype_color, main = 'ref_annotation') 
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_umap_harmony_popv", var = "popv_knn_on_harmony_prediction", opacity = .5, do.label = T, labels.size = 3, color.panel = celltype_color)
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_umap_harmony_popv", var = "popv_knn_on_harmony_prediction_probabilities", opacity = .5, do.label = T, labels.size = 3)

cat('\n\n### Scanorma \n')
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'ref'], reduction.use = "X_umap_scanorma_popv", var = "popv_parent", opacity = .1,
             do.label = T, labels.size = 3, color.panel = celltype_color, main = 'ref_annotation') 
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_umap_scanorma_popv", var = "popv_knn_on_scanorama_prediction", opacity = .5, do.label = T, labels.size = 3, color.panel = celltype_color)
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_umap_scanorma_popv", var = "popv_knn_on_scanorama_prediction_probabilities", opacity = .5, do.label = T, labels.size = 3)

cat('\n\n### BBKNN \n')
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'ref'], reduction.use = "X_bbknn_umap_popv", var = "popv_parent", opacity = .1,
             do.label = T, labels.size = 3, color.panel = celltype_color, main = 'ref_annotation') 
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_bbknn_umap_popv", var = "popv_knn_on_bbknn_prediction", opacity = .5, do.label = T, labels.size = 3, color.panel = celltype_color)
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_bbknn_umap_popv", var = "popv_knn_on_bbknn_prediction_probabilities", opacity = .5, do.label = T, labels.size = 3)

cat('\n\n### CellTypist \n')
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'ref'], reduction.use = "X_scanvi_umap_popv", var = "popv_parent", opacity = .1,
             do.label = T, labels.size = 3, color.panel = celltype_color, main = 'ref_annotation') 
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scanvi_umap_popv", var = "popv_celltypist_prediction", opacity = .5, do.label = T, labels.size = 3, color.panel = celltype_color)
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scanvi_umap_popv", var = "popv_celltypist_prediction_probabilities", opacity = .5, do.label = T, labels.size = 3)

cat('\n\n### SVM \n')
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'ref'], reduction.use = "X_scanvi_umap_popv", var = "popv_parent", opacity = .1,
             do.label = T, labels.size = 3, color.panel = celltype_color, main = 'ref_annotation') 
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scanvi_umap_popv", var = "popv_svm_prediction", opacity = .5, do.label = T, labels.size = 3, color.panel = celltype_color)
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scanvi_umap_popv", var = "popv_svm_prediction_probabilities", opacity = .5, do.label = T, labels.size = 3)


cat('\n\n### RF \n')
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'ref'], reduction.use = "X_scanvi_umap_popv", var = "popv_parent", opacity = .1,
             do.label = T, labels.size = 3, color.panel = celltype_color, main = 'ref_annotation') 
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scanvi_umap_popv", var = "popv_rf_prediction", opacity = .5, do.label = T, labels.size = 3, color.panel = celltype_color)
dittoDimPlot(popv_sce[,popv_sce$X_dataset == 'query'], reduction.use = "X_scanvi_umap_popv", var = "popv_rf_prediction_probabilities", opacity = .5, do.label = T, labels.size = 3)
```

## Outputs
This analysis generated the following results: <br>

1. `AnnData` object with both reference and query datasets containing PopV prediction. <br>
  The object was saved at "`r print(py$out_path)`" <br>
2. `SingleCellExperiment` object with both reference and query datasets containing PopV prediction. <br>
  The object was saved at "`r print(gsub('.h5ad', '_sce.rds', py$out_path))`" <br>


## Session information

```{r session}
etime <- proc.time() - ptm
cat('This analysis took ', etime[3], ' s \n')

sessioninfo::session_info()

```

