---
title: "Validate PopV annotation"
author: "Alex Matei (adapted from Yi-Nan Li)"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    wrap: 72
    code_folding: hide
---

# Setup

```{r setup, include=FALSE}
if (!requireNamespace("here", quietly = TRUE)) {
  install.packages("here")
}
library(here)
#source(here("config.R"))

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, root.dir = here::here())
```

```{r ts}
# setup timestamp
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()
```

This document was created with the time stamp *`r ts`*. <br>

# Load data - **USER INPUT REQUIRED**

Load the annotated data from Alex with BANKSY clustering and Honglin skin dataset from `20240816_CosMxUAP1SSc_FibPopV_HonglinJID_inference.html` and the PopV annotation with the timestamp ‘20240816091252’.


```{r data, message=FALSE}
library(SpatialExperiment)
library(tidyverse)

spe <- readr::read_rds("/mnt/data/home/yinanli/projects/2024_CosMxUAP/data/20241130_CosMxSScSkin_CnMerg_241201111305.rds")
popv_sce <- readr::read_rds("/mnt/data/home/yinanli/projects/2024_CosMxUAP/data/CosMxUAP1SSc_Fib_HonglinJID_popv_20240816091252_sce.rds")

# Rename annotations CosMx dataset -----------------

## Replace "Fibroblasts" with "FB"
spe$banksy_fib <- gsub("Fibroblasts", "FB", spe$banksy_fib)
# Rename several fibroblasts subsets
spe$banksy_fib <- case_when(spe$banksy_fib=="LGALS1+" ~ "ACKR3+",
                               spe$banksy_fib=="IGFBP5+" ~ "GSN+",
                               spe$banksy_fib=="KLF2+" ~ "DUSP+",
                               .default = spe$banksy_fib)

# Rename annotations reference dataset (GSE138669) ----------------

## scvi annotations 
popv_sce$popv_knn_on_scvi_prediction <- case_when (
  popv_sce$popv_knn_on_scvi_prediction == "WIF1+" ~ "COMP+",
  popv_sce$popv_knn_on_scvi_prediction == "PRSS23+" ~ "PRSS23/SFRP2+",
  popv_sce$popv_knn_on_scvi_prediction == "SFRP4+" ~ "SFRP4/SFRP2+",
  TRUE ~ popv_sce$popv_knn_on_scvi_prediction
)

unique(popv_sce$popv_knn_on_scvi_prediction)

## scanvi annotations 
popv_sce$popv_scanvi_prediction <- case_when (
  popv_sce$popv_scanvi_prediction == "WIF1+" ~ "COMP+",
  popv_sce$popv_scanvi_prediction == "PRSS23+" ~ "PRSS23/SFRP2+",
  popv_sce$popv_scanvi_prediction == "SFRP4+" ~ "SFRP4/SFRP2+",
  TRUE ~ popv_sce$popv_scanvi_prediction
)

unique(popv_sce$popv_scanvi_prediction)

# Remove cells that were previously wrongly assigned as fibroblasts from the sce ----------
dim(popv_sce)
popv_sce <- popv_sce[, c(colnames(spe[, spe$lv2.anno_bank == "Fibroblasts"]), colnames(popv_sce[, popv_sce$X_dataset == 'ref']))]
dim(popv_sce)

```

# Visualize annotation {.tabset .tabset-pill}

```{r viz, results='asis'}
library(dittoSeq)
cat('\n\n### Embeddings \n')
dittoDimPlot(
  popv_sce,
  reduction.use = "X_scvi_umap_popv",
  var = "X_dataset",
  opacity = .5,
  do.label = F,
  labels.size = 2,
  main = 'scVI',
  split.by = 'X_dataset',
  show.others = FALSE
)
dittoDimPlot(
  popv_sce,
  reduction.use = "X_scanvi_umap_popv",
  var = "X_dataset",
  opacity = .5,
  do.label = F,
  labels.size = 2,
  main = 'scANVI',
  split.by = 'X_dataset',
  show.others = FALSE
)
# dittoDimPlot(popv_sce, reduction.use = "X_umap_harmony_popv", var = "X_dataset", opacity = .5,
#              do.label = F, labels.size = 2, main = 'Harmony', split.by = 'X_dataset', show.others = FALSE)
# dittoDimPlot(popv_sce, reduction.use = "X_umap_scanorma_popv", var = "X_dataset", opacity = .5,
#              do.label = F, labels.size = 2, main = 'Scanorma', split.by = 'X_dataset', show.others = FALSE)
# dittoDimPlot(popv_sce, reduction.use = "X_bbknn_umap_popv", var = "X_dataset", opacity = .5,
#              do.label = F, labels.size = 2, main = 'BBKNN', split.by = 'X_dataset', show.others = FALSE)

popv_ref_sce <- popv_sce[, popv_sce$X_dataset == 'ref']
popv_ref_sce_sub <- popv_ref_sce[, sample(colnames(popv_ref_sce), 4000, replace = FALSE)]
popv_cosmx_sce <- popv_sce[, popv_sce$X_dataset == 'query']

match_fib_id <- match(colnames(popv_cosmx_sce), colnames(spe[,spe$lv2.anno_bank=="Fibroblasts"]))
popv_cosmx_sce$banksy_fib <- NA
popv_cosmx_sce$banksy_fib[match_fib_id] <- spe[,spe$lv2.anno_bank=="Fibroblasts"]$banksy_fib

cat('\n\n### scVI \n')
dittoDimPlot(
  popv_cosmx_sce,
  reduction.use = "X_scvi_umap_popv",
  var = "banksy_fib",
  opacity = .5,
  do.label = T,
  labels.size = 3,
  main = 'scVI_CosMx',
  legend.show = T,
  size = 1
)
dittoDimPlot(
  popv_cosmx_sce,
  reduction.use = "X_scvi_umap_popv",
  var = "popv_knn_on_scvi_prediction",
  opacity = .5,
  do.label = T,
  labels.size = 3,
  main = 'scVI_CosMx',
  legend.show = T,
  size = 1
)
dittoDimPlot(
  popv_ref_sce_sub,
  reduction.use = "X_scvi_umap_popv",
  var = "popv_knn_on_scvi_prediction",
  opacity = .5,
  do.label = T,
  labels.size = 3,
  main = 'scVI_reference',
  legend.show = T,
  size = 1
)

cat('\n\n### scANVI \n')
dittoDimPlot(
  popv_cosmx_sce,
  reduction.use = "X_scanvi_umap_popv",
  var = "banksy_fib",
  opacity = .5,
  do.label = T,
  labels.size = 3,
  main = 'scANVI_CosMx',
  legend.show = T,
  size = 1
)
dittoDimPlot(
  popv_cosmx_sce,
  reduction.use = "X_scanvi_umap_popv",
  var = "popv_scanvi_prediction",
  opacity = .5,
  do.label = T,
  labels.size = 3,
  main = 'scANVI_CosMx',
  legend.show = T,
  size = 1
)
dittoDimPlot(
  popv_ref_sce_sub,
  reduction.use = "X_scanvi_umap_popv",
  var = "popv_scanvi_prediction",
  opacity = .5,
  do.label = T,
  labels.size = 3,
  main = 'scANVI_reference',
  legend.show = T,
  size = 1
)

#cat('\n\n### Harmony \n')
# dittoDimPlot(popv_cosmx_sce, reduction.use = "X_umap_harmony_popv", var = "popv_knn_on_harmony_prediction", opacity = .5,
#              do.label = T, labels.size = 3, main = 'Harmony_cosmx', legend.show = T, size = 1)
# dittoDimPlot(popv_ref_sce_sub, reduction.use = "X_umap_harmony_popv", var = "popv_knn_on_harmony_prediction", opacity = .5,
#              do.label = T, labels.size = 3, main = 'Harmony_gse249279', legend.show = T, size = 1) 
```



# Visualize annotations embeddings figure

```{r prep viz anno emb fig}

reducedDimNames(popv_sce) <- case_when(
  reducedDimNames(popv_sce) == "X_scanvi_umap_popv" ~ "scANVI_UMAP",
  TRUE ~ reducedDimNames(popv_sce)
)
popv_sce$refquery <- as.character(popv_sce$X_dataset)
popv_sce$refquery <- case_when(popv_sce$refquery == "ref" ~ "GSE138669",
                               popv_sce$refquery == "query" ~ "cISH")
popv_sce$refquery <- factor(popv_sce$refquery, levels = c(unique(popv_sce$refquery)[2], unique(popv_sce$refquery)[1]))

```


```{r viz anno emb fig, fig.dim=c(3,3)}

dittoDimPlot(
  popv_sce,
  reduction.use = "scANVI_UMAP",
  var = "refquery",
  size = 2,
  opacity = .5,
  do.label = F,
  labels.size = 4,
  main = NULL,
  split.by = 'refquery',
  legend.show = FALSE,
  show.others = FALSE
) +
  theme(strip.text = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15))

```

# Vizualize annotations label transfer figure

```{r prep viz anno lab tr fig}

reducedDimNames(popv_cosmx_sce) <- case_when(
  reducedDimNames(popv_cosmx_sce) == "X_scanvi_umap_popv" ~ "scANVI_UMAP",
  TRUE ~ reducedDimNames(popv_cosmx_sce)
)

reducedDimNames(popv_ref_sce_sub) <- case_when(
  reducedDimNames(popv_ref_sce_sub) == "X_scanvi_umap_popv" ~ "scANVI_UMAP",
  TRUE ~ reducedDimNames(popv_ref_sce_sub)
)

```

```{r viz anno lab tr fig, fig.width=10, fig.height=4}
library(patchwork)

p1<-dittoDimPlot(
  popv_cosmx_sce,
  reduction.use = "scANVI_UMAP",
  var = "banksy_fib",
  opacity = .5,
  do.label = T,
  labels.size = 3,
  main = 'scANVI_CosMx',
  legend.show = T,
  size = 2
)
p2<-dittoDimPlot(
  popv_cosmx_sce,
  reduction.use = "scANVI_UMAP",
  var = "popv_scanvi_prediction",
  opacity = .5,
  do.label = T,
  labels.size = 3,
  main = 'scANVI_CosMx',
  legend.show = T,
  size = 2
)
p3<-dittoDimPlot(
  popv_ref_sce_sub,
  reduction.use = "scANVI_UMAP",
  var = "popv_scanvi_prediction",
  opacity = .5,
  do.label = T,
  labels.size = 3,
  main = 'scANVI_reference',
  legend.show = F,
  size = 2
)

p3|p2

```


# Comparison with previous annotation (Clustering/BANKSY) {.tabset .tabset-pill}

```{r cp-anno, results='asis'}
library(dplyr)
library(stringr)
scanvi_anno <- popv_cosmx_sce$popv_scanvi_prediction %>% `names<-`(colnames(popv_cosmx_sce))
scvi_anno <- popv_cosmx_sce$popv_knn_on_scvi_prediction %>% `names<-`(colnames(popv_cosmx_sce))
# harmony_anno <- popv_cosmx_sce$popv_knn_on_harmony_prediction %>% `names<-`(colnames(popv_cosmx_sce))

spe_sub <- spe[,spe$lv2.anno_bank == "Fibroblasts"]
scanvi_anno <- scanvi_anno[colnames(spe_sub)]
scvi_anno <- scvi_anno[colnames(spe_sub)]
# harmony_anno <- harmony_anno[colnames(spe_sub)]

spe_sub$popv_scanvi <- scanvi_anno
spe_sub$popv_scvi <- scvi_anno
# spe_sub$popv_harmony <- harmony_anno


plot_conti_ht <- function(tbl, xlab, ylab) {
  tmp <- reshape2::melt(tbl)
  p <- ggplot(tmp, aes(x = Var2, y = Var1, fill = value)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "blue") +
    labs(x = xlab, y = ylab, fill = "prop") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  return(p)
}


banksy_lv3_scvi <- prop.table(table(spe_sub$popv_scvi, spe_sub$banksy_fib), 1) 
banksy_lv3_scanvi <- prop.table(table(spe_sub$popv_scanvi, spe_sub$banksy_fib), 1) 
# banksy_lv3_harmony <- prop.table(table(spe_sub$popv_harmony, spe_sub$banksy_fib), 1) 

banksy_lv3_scvi2 <- prop.table(table(spe_sub$popv_scvi, spe_sub$banksy_fib), 2) 
banksy_lv3_scanvi2 <- prop.table(table(spe_sub$popv_scanvi, spe_sub$banksy_fib), 2) 
# banksy_lv3_harmony2 <- prop.table(table(spe_sub$popv_harmony, spe_sub$banksy_fib), 2)

cat('\n\n### scANVI and BANKSY ref in query \n')
plot_conti_ht(banksy_lv3_scanvi, 'BANKSY_lv3', 'PopV_scANVI')

cat('\n\n### scANVI and BANKSY query in ref \n')
plot_conti_ht(banksy_lv3_scanvi2, 'BANKSY_lv3', 'PopV_scANVI')

cat('\n\n### scVI and BANKSY ref in query \n')
plot_conti_ht(banksy_lv3_scvi, 'BANKSY_lv3', 'PopV_scVI')

cat('\n\n### scVI and BANKSY query in ref \n')
plot_conti_ht(banksy_lv3_scvi2, 'BANKSY_lv3', 'PopV_scVI')

# cat('\n\n### Harmony and BANKSY \n')
# plot_conti_ht(banksy_lv3_harmony, 'BANKSY_lv3', 'PopV_Harmony')

# Var2 <- case_when(endsWith(spe_sub$lv3.anno, 'Fib') ~ NA, .default = spe_sub$lv3.anno)
# 
# anno_lv3_scvi <- prop.table(table(spe_sub$popv_scvi, Var2), 1) 
# anno_lv3_scanvi <- prop.table(table(spe_sub$popv_scanvi, Var2), 1) 
# anno_lv3_harmony <- prop.table(table(spe_sub$popv_harmony, Var2), 1)
# 
# cat('\n\n### scVI and direct clustering \n')
# plot_conti_ht(anno_lv3_scvi, 'anno_lv3', 'PopV_scANVI')
# cat('\n\n### scANVI and direct clustering \n')
# plot_conti_ht(anno_lv3_scanvi, 'anno_lv3', 'PopV_scVI')
# cat('\n\n### Harmony and direct clustering \n')
# plot_conti_ht(anno_lv3_harmony, 'anno_lv3', 'PopV_Harmony')


```

# Sanky plots

```{r sanky}
library(ggsankey)

# Create data frames
sanky_df_scvi <- data.frame(banksy=spe_sub$banksy_fib, scvi=spe_sub$popv_scvi)
sanky_df_scanvi <- data.frame(banksy=spe_sub$banksy_fib, scanvi=spe_sub$popv_scanvi)

matching <- sanky_df_scanvi

# Summarize and calculate frequency
matching.summary <- matching %>%
  group_by(across(all_of(colnames(matching)))) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(across(all_of(colnames(matching)[2]))) %>%
  mutate(tot_cells = sum(count), freq = count / tot_cells) %>%
  ungroup()

# Identify rare matches (freq < 0.1 or count < 20)
rare.matches <- matching.summary %>%
  filter(freq < 0.1 | count < 20) %>%
  select(all_of(colnames(matching[1:2])))

# Remove rare matches from the original data
matching.filtered <- matching %>%
  anti_join(rare.matches, by = colnames(matching)[1:2])

# Create sankey data with filtered matches
matching.sankey <- make_long(matching.filtered, !!sym(colnames(matching)[1]), !!sym(colnames(matching)[2]))

# Plot the simplified Sankey
ggplot(
  matching.sankey,
  aes(
    x = x,
    next_x = next_x,
    node = node,
    next_node = next_node,
    fill = factor(node),
    label = node
  )
) +
  geom_sankey(flow.alpha = 0.3, node.color = "gray30") +
  geom_sankey_label(size = 3,
                    color = "white",
                    fill = "gray40") +
  theme_sankey(base_size = 18) +
  labs(x = NULL) +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))


```


# Session information

```{r session}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```

```{bash, eval=FALSE}
Rscript -e "rmarkdown::render('/mnt/data/home/alexmatei/CosMx_fistrun_Analysis_AEM/Analysis/PopV annotation/20241207_validate_popv.Rmd', output_file = 'Plots_validation_PopV_anno_Honglin.html', output_dir = '/mnt/data/home/alexmatei/CosMx_fistrun_Analysis_AEM/Results/PopV_Annotation/')" > /mnt/data/home/alexmatei/CosMx_fistrun_Analysis_AEM/Logs/log_$(date +"%Y%m%d_%H%M%S").log 2>&1

```

